\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-30 docs/parco13/paper.tex
Work through section 3 in paper

Mesh modification directives, such as adaptive refinement and coarsening, are
called from \forestclaw and executed as black-box operations inside \pforest.
The only information that flows back to \pforest is a set of per-patch
refinement and coarsening flags, which are computed from the numerical state in
\forestclaw.  Our current algorithm requires neighbor patches to have a size
difference of at most a factor of two (the 2:1 balance condition).  This condition
can have nonlocal effects on the refinement pattern and is enforced by
special-purpose parallel algorithms in \pforest \cite{IsaacBursteddeGhattas12}.

@@

\forestclaw uses iterators over all rank-local leaves, optionally restricted to
a given level.  Random access is possible and used when executing $\cO(1)$ time
neighbor lookups.  Looping over the patches in the order prescribed by the
forest
\alert{Review 2 - Explain the preceding paragraph in more detail}
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

\alert{The exchange of ghost} data and parallel repartitioning is
provided transparently by \pforest.
\alert{Review 2 requests more info on how this is done.  Sounds like they are more curious about the refinement overhead to other leaves on the tree and the strategy as to how refinement is decided.
%
Quote: How is AMR working in detail. For example: Figure 1 shows a
2:1 refinement, thus refinement of a single patch could affect a whole region
of the computational domain. + The underlying equations are known to
Forestclaw: How is is the information shared with p4est (equation dependent
refinement criteria, refinement and coarsening strategies (interpolation,
integration).
}

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, the time step length, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  When using sub-cycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

@@

We provide two examples that demonstrate the \forestclaw code.
The numerical results to date have been designed to verify that
interface layer between \pforest and \forestclaw is sufficiently
flexible and robust.  Of particular importance was ensuring that all
ghost cell transfers (including averaging and
interpolation) have been implemented correctly.
The basic \clawpack algorithm and corresponding code have been
thoroughly tested and need no further verification in our context.

>>>>

Mesh modification directives, such as adaptive refinement and coarsening, are
called from \forestclaw and executed as black-box operations inside \pforest.
The only information that flows into \pforest is a set of per-patch
refinement and coarsening flags, which are computed from the numerical state in
\forestclaw.  Our current algorithm requires neighbor patches to have a size
difference of at most a factor of two (the 2:1 balance condition).  This condition
can have nonlocal effects on the refinement pattern and is enforced by
special-purpose parallel algorithms in \pforest \cite{IsaacBursteddeGhattas12}.

@@

\forestclaw uses iterators over all rank-local leaves to execute numerical
tasks one patch at a time, optionally restricted to a given refinement level.
Random access to patches is possible and used when dereferencing the results
from
% executing
% $\cO(1)$ time
neighbor lookups.  Looping over the patches in the order
prescribed by the forest
% \alert{Review 2 - Explain the preceding paragraph in more detail}
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.
% Parallel neighbor exchanges are hidden
% inside \pforest and called by \forestclaw
% % at well-defined synchronization points
% as follows.

With MPI parallelism, neighbor patches can be either local or assigned to a
different process (ghost patches).  Since this must make no difference
numerically, we need to ensure that the values of ghost patches are current
whenever they are accessed.  To this end, we allocate storage for a layer of
ghost patches in \forestclaw, which we can pass to a general-purpose \pforest
routine that communicates local leaf data to all processes that view a
particular leaf as a ghost.  If we call this routine before we go through the
local neighbor interactions, we can handle ghost values implicitly by neighbor
lookups without querying if they are local or remote.

% \alert{The exchange of ghost} data and parallel repartitioning is
% provided transparently by \pforest.
% \alert{Review 2 requests more info on how this is done.  Sounds like they are more curious about the refinement overhead to other leaves on the tree and the strategy as to how refinement is decided.
% %
% Quote: How is AMR working in detail. For example: Figure 1 shows a
% 2:1 refinement, thus refinement of a single patch could affect a whole region
% of the computational domain. + The underlying equations are known to
% Forestclaw: How is is the information shared with p4est (equation dependent
% refinement criteria, refinement and coarsening strategies (interpolation,
% integration).
% }
% CB: Is this addressed sufficiently now?

% count of exchanges with and without sub-cycling
In the context of finite-element or finite-difference methods, there should be
one such parallel data exchange per time step for a global value of the time
step length $\dt$, or one exchange per discretization level per time step if
$\dt$ is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  In the tradition of block-structured AMR codes, interaction
between neighbor patches is done in a hierarchy from coarse to fine levels, and
then correction factors are propagated back from fine to coarse levels, requiring
an exchange at each level.  With sub-cycling, this entails a recursion with an
operation count that is exponential in the difference between the largest and
the smallest refinement level, with the benefit that $\dt$ matches the CFL
condition at each level.  While a large number of exchanges per time step can
present a scaling bottleneck due to the inherent synchronization and latency
losses, the expectation is still that the time to solution improves when
switching from uniform to adaptive meshes, since we are computing with fewer
patches, and improves again when enabling sub-cycling, since we take larger
time steps on the coarser levels.

% load balance / (weighted) partition
The space-filling curve paradigm allows for lightweight repartitioning
algorithms.  Even uniform meshes benefit from this approach since the number of
processors does not need to be commensurable with the number of patches in each
space dimension.  For adaptive meshes, we repartiton after every refinement
operation and transfer the numerical data accordingly; see \cite[Fig.\
4]{BursteddeGhattasStadlerEtAl08} for the overall procedure.  When using
sub-cycling, we have the option to assign a weight to each patch based on its
level, determined by the expected number of sub-cycles per coarse-level time
step, and partition to equidistribute the cumulative weight between the
processors.  Currently, we are not enforcing a load balance that is attained
separately for each level, as it is done for example in Chombo
\cite{ColellaGravesKeenEtAl07} or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

@@

We provide two examples that demonstrate the \forestclaw code.
The numerical results to date have been designed to verify that
interface layer between \pforest and \forestclaw is sufficiently
flexible and robust.  Of particular importance was ensuring that all
ghost cell transfers (including averaging and
interpolation) are implemented correctly.
The basic \clawpack algorithm and corresponding code are
thoroughly tested and need no further verification in our context.
